pema application

### 2022-04-07 paper/application.Rmd
application

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages
First, load in the necessary packages. In addition, if you are running the model locally on a multicore machine, you can set `options(mc.cores = 4)`. This ensures the different MCMC chains will be run in parallel, making estimation faster.

```{r}
library(pema)
options(mc.cores = 4)
library(tidyverse)
library(fastDummies)
```

@@

Q1 Caspar: de data lijkt nog niet beschikbaar in de laatste versie van het package?

```{r data}
bonapersona <- read.csv2("./pema_application/MAB_data.csv")
summary(bonapersona)
bonapersona$yi <- as.numeric(bonapersona$yi)
bonapersona$vi <- as.numeric(bonapersona$vi)
bonapersona$bias <- as.numeric(bonapersona$bias)
```
For this application, we use the same 26 moderators as in Bonapersona et al. (2019; see Figure 3A for an overview). We select these moderators and create dummy variables for all categorical moderators.

```{r dummies}
datsel <- bonapersona %>% 
  select(yi, vi, author, mTimeLength, year, model, ageWeek, testAuthorGrouped,
         strainGrouped, bias, species, domain, mLDGrouped, waterTcate, 
         origin, mCageGrouped, seqGeneration, baseline, blindExp,
         retentionGrouped, hit2Grouped, outBlind, sex, housing,
         testLDGrouped, freezingType, blindRand, control, mControlGrouped)

# Recode categorical moderators
catmod <- c("model", "testAuthorGrouped", "strainGrouped", "species", "domain", "mLDGrouped", "waterTcate", "origin", 
            "mCageGrouped", "seqGeneration", "baseline", "blindExp", "retentionGrouped", "hit2Grouped",
            "outBlind", "sex", "housing", "testLDGrouped", "freezingType", "blindRand", "control", "mControlGrouped")
datdum <- dummy_cols(datsel, 
                     select_columns = catmod,
                     remove_first_dummy = TRUE,
                     remove_selected_columns = TRUE) 
```

## Two-level model
First, for simplicity, we run a two-level model ignoring the fact that certain effect sizes come from the same study.

```{r}
dat2l <- datdum %>%
  select(-author)
```

### Two-level model with the lasso prior
We start with running a penalized meta-analysis using the lasso prior. Compared to the horseshoe prior, the lasso is easier to use because it has only two hyperparameters to set. However, the lighter tails of the lasso can result in large coefficients being shrunken too much towards zero thereby leading to potentially more bias compared to the regularized horseshoe prior.

@@

Q2 Caspar: `standardize = TRUE` resulteert in NAs waardoor Stan een error geeft. Weet jij hoe dit kan? Daarnaast moeten we misschien een meer informatieve waarschuwing geven in het algemeen als er NAs in de data zitten?

```{r}
# Manually standardize the continuous moderators, because `standardize = TRUE` does not work for some reason
dat2l <- dat2l %>%
  mutate(year = scale(year)) %>%
  mutate(ageWeek = scale(ageWeek)) %>%
  mutate(bias = scale(bias))
```

```{r lasso, cache = TRUE}
fit.lasso <- brma(yi ~ .,
            data = dat2l,
            vi = "vi",
            method = "lasso",
            standardize = FALSE,
            prior = c(df = 1, scale = 1),
            mute_stan = FALSE)
```

#### Assessing convergence and interpreting the results
We can request the results using the `summary` function. Before we interpret the results, we need to ensure that the MCMC chains have converged to the posterior distribution. Two helpful diagnostics provided in the summary are the number of effective posterior samples `n_eff` and the potential scale reduction factor `Rhat`. `n_eff` is an estimate of the number of independent samples from the posterior. Ideally, the ratio `n_eff` to total samples is as close to 1 as possible. `Rhat` compares the between- and within-chain estimates and is ideally close to 1 (indicating the chains have mixed well). Should any values for `n_eff` or `Rhat` be far from these ideal values, you can try increasing the number of iterations through the `iter` argument. By default, the `brma` function runs four MCMC chains with 2000 iterations each, half of which is discarded as burn-in. As a result, a total of 4000 iterations is available on which posterior summaries are based. If this does not help, non-convergence might indicate a problem with the model specification.

```{r}
options(max.print = 10000)
summary(fit.lasso)
```

@@

Q3 Caspar: Deze resultaten lijken niet overeen te komen met die uit het originele paper waar metaforest is gebruikt. Daar lijkt sex namelijk niet zo'n belangrijke moderator (gebaseerd op figuur 3A). Ik weet alleen niet genoeg van metaforest om te bepalen of dit problematisch is of logisch te verklaren door de onderliggende methode.

### Two-level model with the horseshoe prior
Next, we look into the regularized horseshoe prior. The horseshoe prior has five hyperparameters that can be set. Three parameters are degrees of freedom parameters which influence the tails of the distributions in the prior. Generally, it is not needed to specify different values for these hyperparameters. Here, we focus instead on the global scale parameter and the scale of the slab.

```{r hs, cache = TRUE}
# use the default settings
fit.hs1 <- brma(yi ~ .,
               data = dat2l,
               vi = "vi",
               method = "hs",
               standardize = FALSE,
               prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = NULL),
               mute_stan = FALSE)

# reduce the global scale
fit.hs2 <- brma(yi ~ .,
                data = dat2l,
                vi = "vi",
                method = "hs",
                standardize = FALSE,
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 0.1, scale_slab = 1, relevant_pars = NULL),
                mute_stan = FALSE)

# increase the scale of the slab
fit.hs3 <- brma(yi ~ .,
                data = dat2l,
                vi = "vi",
                method = "hs",
                standardize = FALSE,
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 5, relevant_pars = NULL),
                mute_stan = FALSE)

@@

Note that the horseshoe prior results in some divergent transitions. This can be an indication of non-convergence. However, these divergences arise often when using the horseshoe prior and as long as there are not too many of them, the results can still be used.

@@

```{r}
df.fun <- function(fit, prior){
  plotdat <- data.frame(fit$coefficients)
  plotdat$par <- rownames(plotdat)
  plotdat$Prior <- prior
  return(plotdat)
}

df0 <- df.fun(fit.lasso, prior = "lasso")
df1 <- df.fun(fit.hs1, prior = "hs default")
df2 <- df.fun(fit.hs2, prior = "hs reduced global scale")
df3 <- df.fun(fit.hs3, prior = "hs increased slab scale")

@@

# select pars with highest posterior mean estimates 
parsel <- df1 %>%
  arrange(desc(abs(`mean`))) %>%
  select(par) %>%
  filter(!par == "tau2")

@@

ggplot(dfsub, aes(x=mean, y=par, group = Prior)) + 
  geom_errorbar(aes(xmin=X2.5., xmax=X97.5., colour = Prior), width=.1, position = position_dodge(width = pd)) +
  geom_point(aes(colour = Prior), position = position_dodge(width = pd)) +
  theme_bw() + xlab("Posterior mean") + ylab("")
```

@@

## Three-level model
Finally, we can also take into account the fact that some effect sizes might come from the same study by fitting a three-level model as follows:

```{r 3level, cache = TRUE}
fit.3l <- brma(yi ~ .,
               data = datdum,
               vi = "vi",
               study = "author",
               method = "lasso",
               standardize = FALSE,
               prior = c(df = 1, scale = 1),
               mute_stan = FALSE)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
run_everything = FALSE
```

## Packages

First, load in the necessary packages. In addition, if you are running the model locally on a multicore machine, you can set `options(mc.cores = 4)`. This ensures the different MCMC chains will be run in parallel, making estimation faster.

```{r}
library(pema)
options(mc.cores = 4)
library(tidyverse)
library(fastDummies)
library(tidySEM)
library(DT)
```

@@

```{r data}
tab_desc <- descriptives(bonapersona)[c("name", "type", "n", "unique", "mean", "sd", "v")]
DT::datatable(tab_desc) %>%
  formatRound(columns=c("mean", "sd", "v"), digits = 2)
```

## Impute missings

```{r}
bonapersona$ageWeek[is.na(bonapersona$ageWeek)] <- median(bonapersona$ageWeek, na.rm = TRUE)
```

## Moderators

For this application, we use a smaller selection of moderators than in Bonapersona et al. (2019; see Figure 3A for an overview).

```{r dummies}
datsel <- bonapersona %>% 
  select(yi, vi, author, mTimeLength, year, model, ageWeek, strainGrouped, bias, species, domain, sex) 
```

## Two-level model

First, for simplicity, we run a two-level model ignoring the fact that certain effect sizes come from the same study.

```{r}
dat2l <- datsel %>%
  select(-author)
```

### Two-level model with the lasso prior

We start with running a penalized meta-analysis using the lasso prior. Compared to the horseshoe prior, the lasso is easier to use because it has only two hyperparameters to set. However, the lighter tails of the lasso can result in large coefficients being shrunken too much towards zero thereby leading to potentially more bias compared to the regularized horseshoe prior.

@@

```{r, echo = TRUE, eval = F}
fit_lasso <- brma(yi ~ .,
                  data = dat2l,
                  vi = "vi",
                  method = "lasso",
                  prior = c(df = 1, scale = 1),
                  mute_stan = FALSE)
```
```{r, echo = FALSE, eval = TRUE}
if(run_everything){
  fit_lasso <- brma(yi ~ .,
                    data = dat2l,
                    vi = "vi",
                    method = "lasso",
                    prior = c(df = 1, scale = 1),
                    mute_stan = FALSE)
  saveRDS(fit_lasso, "fitlasso.RData")
}
fit_lasso <- readRDS("fitlasso.RData")
```

#### Assessing convergence and interpreting the results

We can request the results using the `summary` function. Before we interpret the results, we need to ensure that the MCMC chains have converged to the posterior distribution. Two helpful diagnostics provided in the summary are the number of effective posterior samples `n_eff` and the potential scale reduction factor `Rhat`. `n_eff` is an estimate of the number of independent samples from the posterior. Ideally, the ratio `n_eff` to total samples is as close to 1 as possible. `Rhat` compares the between- and within-chain estimates and is ideally close to 1 (indicating the chains have mixed well). Should any values for `n_eff` or `Rhat` be far from these ideal values, you can try increasing the number of iterations through the `iter` argument. By default, the `brma` function runs four MCMC chains with 2000 iterations each, half of which is discarded as burn-in. As a result, a total of 4000 iterations is available on which posterior summaries are based. If this does not help, non-convergence might indicate a problem with the model specification.

```{r, eval = TRUE}
sum <- summary(fit_lasso)
DT::datatable(sum$coefficients[, c("mean", "sd", "2.5%", "97.5%", "n_eff", "Rhat")]) %>%
  formatRound(columns=c("mean", "sd", "2.5%", "97.5%", "n_eff", "Rhat"), digits = 2)
```

@@

Also note the summary statistics for `tau2`, the (unexplained) residual between-studies heterogeneity.
The 95% credible interval for this coefficient excludes zero, indicating that there is a non-zero amount of unexplained heterogeneity.
It is customary to express this heterogeneity in terms of $I^2$, the percentage of variation across studies that is due to heterogeneity rather than chance (Higgins and Thompson, 2002; Higgins et al., 2003).
The helper function `I2()` computes the posterior distribution of $I^2$ based on the MCMC draws of `tau2`:

```{r}
I2(fit_lasso)
```

### Two-level model with the horseshoe prior

Next, we look into the regularized horseshoe prior. The horseshoe prior has five hyperparameters that can be set. Three parameters are degrees of freedom parameters which influence the tails of the distributions in the prior. Generally, it is not needed to specify different values for these hyperparameters. Here, we focus instead on the global scale parameter and the scale of the slab.

```{r hs, echo = TRUE, eval = F}
# use the default settings
fit_hs1 <- brma(yi ~ .,
                data = dat2l,
                vi = "vi",
                method = "hs",
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = NULL),
                mute_stan = FALSE)

# reduce the global scale
fit_hs2 <- brma(yi ~ .,
                data = dat2l,
                vi = "vi",
                method = "hs",
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 0.1, scale_slab = 1, relevant_pars = NULL),
                mute_stan = FALSE)

# increase the scale of the slab
fit_hs3 <- brma(yi ~ .,
                data = dat2l,
                vi = "vi",
                method = "hs",
                prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 5, relevant_pars = NULL),
                mute_stan = FALSE)

@@

```{r, echo = FALSE, eval = TRUE}
if(run_everything){
  # use the default settings
  fit_hs1 <- brma(yi ~ .,
                  data = dat2l,
                  vi = "vi",
                  method = "hs",
                  prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 1, relevant_pars = NULL),
                  mute_stan = FALSE)

  # reduce the global scale
  fit_hs2 <- brma(yi ~ .,
                  data = dat2l,
                  vi = "vi",
                  method = "hs",
                  prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 0.1, scale_slab = 1, relevant_pars = NULL),
                  mute_stan = FALSE)

  # increase the scale of the slab
  fit_hs3 <- brma(yi ~ .,
                  data = dat2l,
                  vi = "vi",
                  method = "hs",
                  prior = c(df = 1, df_global = 1, df_slab = 4, scale_global = 1, scale_slab = 5, relevant_pars = NULL),
                  mute_stan = FALSE)
  saveRDS(fit_hs1, "fit_hs1.RData")
  saveRDS(fit_hs2, "fit_hs2.RData")
  saveRDS(fit_hs3, "fit_hs3.RData")
}
fit_hs1 <- readRDS("fit_hs1.RData")
fit_hs2 <- readRDS("fit_hs2.RData")
fit_hs3 <- readRDS("fit_hs3.RData")
```

Note that the horseshoe prior results in some divergent transitions. This can be an indication of non-convergence. However, these divergences arise often when using the horseshoe prior and as long as there are not too many of them, the results can still be used.<!-- CJ Reference? -->

@@

```{r, eval = F}
make_plotdat <- function(fit, prior){
  plotdat <- data.frame(fit$coefficients)
  plotdat$par <- rownames(plotdat)
  plotdat$Prior <- prior
  return(plotdat)
}

df0 <- make_plotdat(fit_lasso, prior = "lasso")
df1 <- make_plotdat(fit_hs1, prior = "hs default")
df2 <- make_plotdat(fit_hs2, prior = "hs reduced global scale")
df3 <- make_plotdat(fit_hs3, prior = "hs increased slab scale")

@@

# select pars with highest posterior mean estimates 
parsel <- df1 %>%
  arrange(desc(abs(`mean`))) %>%
  select(par) %>%
  filter(!par %in% c("Intercept", "tau2"))

@@

ggplot(dfsub, aes(x=mean, y=par, group = Prior)) + 
  geom_errorbar(aes(xmin=X2.5., xmax=X97.5., colour = Prior), width=.1, position = position_dodge(width = pd)) +
  geom_point(aes(colour = Prior), position = position_dodge(width = pd)) +
  geom_vline(xintercept = 0) +
  theme_bw() + xlab("Posterior mean") + ylab("")
```

@@

## Three-level model

Finally, we can also take into account the fact that some effect sizes might come from the same study by fitting a three-level model as follows:

```{r, echo = TRUE, eval = F}
fit_3l <- brma(yi ~ .,
               data = datsel,
               vi = "vi",
               study = "author",
               method = "lasso",
               standardize = FALSE,
               prior = c(df = 1, scale = 1),
               mute_stan = FALSE)
```
```{r, eval = TRUE, echo = F}
if(run_everything){
  fit_3l <- brma(yi ~ .,
                 data = datsel,
                 vi = "vi",
                 study = "author",
                 method = "lasso",
                 prior = c(df = 1, scale = 1),
                 mute_stan = FALSE)
  saveRDS(fit_3l, "fit_3l.RData")
}
fit_3l <- readRDS("fit_3l.RData")
```

## Standardization

It is possible to override the default standardization, which standardizes all variables (including dummies).
To do so, first manually standardize any variables that must be standardized.
Then, in the call to `brma()`, provide a named list with elements `list(center = ..., scale = ...)`.
For variables that are **not** to be standardized, use `center = 0, scale = 1`.
This retains their original scale.
For variables that are to be standardized, use their original `center` and `scale` to restore the coefficients to their original scale.

In the example below, we standardize a continuous predictor, but we do not standardize the dummies:

```{r echo = FALSE, eval = TRUE}

if(run_everything){
  moderators <- model.matrix(yi~ageWeek + strainGrouped, data = datsel)[, -1]
  scale_age <- scale(moderators[,1])
  stdz <- list(center = c(attr(scale_age, "scaled:center"), rep(0, length(levels(datsel$strainGrouped))-1)),
               scale = c(attr(scale_age, "scaled:scale"),   rep(1, length(levels(datsel$strainGrouped))-1)))
  moderators <- data.frame(datsel[c("yi", "vi", "author")], moderators)
  fit_std <- brma(yi ~ .,
                  data = moderators,
                  vi = "vi",
                  study = "author",
                  method = "lasso",
                  prior = c(df = 1, scale = 1),
                  standardize = stdz,
                  mute_stan = FALSE)
  saveRDS(fit_std, "fit_std.RData")
}
fit_std <- readRDS("fit_std.RData")

```{r echo = TRUE, eval = FALSE}
moderators <- model.matrix(yi~ageWeek + strainGrouped, data = datsel)[, -1]
scale_age <- scale(moderators[,1])
stdz <- list(center = c(attr(scale_age, "scaled:center"), rep(0, length(levels(datsel$strainGrouped))-1)),
             scale = c(attr(scale_age, "scaled:scale"),   rep(1, length(levels(datsel$strainGrouped))-1)))
moderators <- data.frame(datsel[c("yi", "vi", "author")], moderators)
fit_std <- brma(yi ~ .,
                data = moderators,
                vi = "vi",
                study = "author",
                method = "lasso",
                prior = c(df = 1, scale = 1),
                standardize = stdz,
                mute_stan = FALSE)

Note that, in this example, only `ageWeek` is standardized; the remaining (dummy) variables are untouched.
The object `stdz` provides the original center and scale for `ageWeek`, which allows `brma()` to properly rescale the coefficient for this moderator.
As the center and scale for all remaining (dummy) variables are 0 and 1, these coefficients are **not** rescaled.
See the `pema` paper for a discussion of standardization and further references.
